---
title: "Course5-Assignment2"
output: html_document
---

Instructions

Introduction

It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a Fitbit, Nike Fuelband, or Jawbone Up. These type of devices are part of the "quantified self" movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. But these data remain under-utilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

The data for this assignment can be downloaded from the course web site:

Dataset: Activity monitoring data [52K]
The variables included in this dataset are:

steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)
date: The date on which the measurement was taken in YYYY-MM-DD format
interval: Identifier for the 5-minute interval in which measurement was taken
The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.

```{r}

# Read in data from file activity.csv
activity <- read.csv("activity.csv", na.strings="NA")

# Exclude NA
activity$steps <- as.numeric(activity$steps)
activity$interval <- as.numeric(activity$interval)

# Calculate Total Number of Steps
StepsTotal <- aggregate(steps ~ date, data=activity, sum, na.rm=TRUE)


```

Plot a histogram of Total Steps Per Day

```{r, echo=FALSE}
hist(x=StepsTotal$steps, col = "red", breaks=40, xlab="Steps Per Day", main="Daily Number of Steps", cex=1, cex.lab=0.95, cex.axis=0.95, cex.main=0.95, cex.sub=0.95, font=2, font.lab=2)
```

Calculate and report Mean and Median for Total Steps Per Day

```{r, echo=FALSE}
summary(StepsTotal$steps)
```

The next part answered the following question.

What is the average daily activity pattern?

A time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis) was constructed.

```{r, echo=FALSE}

library(lattice)
activity$date <- as.character(activity$date)
time_series <- tapply(activity$steps, activity$interval, mean, na.rm = TRUE)
```
```{r, echo=TRUE}
plot(row.names(time_series), time_series, type = "l", xlab = "5 minute Intervals", ylab="Average All Days", main="Average Steps Taken", col="blue", cex=1, cex.axis=0.75, cex.lab=0.75, cex.main=0.95, font.lab=2, font=2)
```


Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r, echo=FALSE}
maximuminterval <- which.max(time_series) #queries time series for max steps
names(maximuminterval) #gives the 5-min interval with max steps
```

Imputing missing values

There are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.

Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r, echo=FALSE}
sum(is.na(activity)) #calculates the total number of missing values
```

The strategy used to fill in the missing values is to use the average # of steps for the 5-minute interval in which a NA appeared.

```{r, echo=FALSE}
activity2 <- subset(activity, !is.na(activity$steps))
dAta <- activity
imPact <- is.na(dAta$steps)
tAvg <- tapply(activity2$steps, activity2$interval, mean, na.rm = TRUE, simplify = T)
dAta$steps[imPact] <- tAvg[as.character(dAta$interval[imPact])]
```

Here's a histogram of the Total Number of Steps per Day with the adjustment for the NAs.
```{r, echo=FALSE}
totalStepsperday2 <- aggregate(steps ~ date, dAta, FUN = sum)
 par(mar = c(5,8,3,1))
```
```{r, echo=TRUE}
hist(totalStepsperday2$steps, breaks = 40, col = "red", xlab = "Steps Per Day", main = "Daily Number of Steps")
```

Here's the Mean and Median for the total steps per day data adjusted for NAs.

```{r, echo=FALSE}
summary(totalStepsperday2$steps)
```

Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.
Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data

```{r, echo=TRUE}
dAta$date <- as.Date(dAta$date)
dAta$day <- weekdays(dAta$date)
dAta$indicate_day <- as.factor(ifelse(dAta$day == "Saturday" | dAta$day == "Sunday", "weekend", "weekday"))

StepsImputedInterval <- aggregate(steps ~ interval + indicate_day, dAta, mean)
library(lattice)
xyplot(StepsImputedInterval$steps ~ StepsImputedInterval$interval | StepsImputedInterval$indicate_day, main="Average Steps (adjusted for NAs) per Day by Interval", xlab="Interval (5 seconds each)", ylab="Imputed Steps", layout=c(1,2), type="l", cex=1, cex.axis=0.75, font=2, font.lab=2, font.main=2, font.sub=2, font.lab=2)
```

